---
title: 'Developing Data Products: Week 2 Assignment'
output: html_document
---
##### *August 8th, 2020*

## Instructions

Create a web page using R Markdown that features a map created with Leaflet. Host your webpage on either GitHub Pages, RPubs, or NeoCities. Your webpage must contain the date that you created the document and it must contain a map created with Leaflet.

## Overview of Map

This map shows the nine Universities of California (Berkeley, Davis, LA, Santa Cruz, Irvine, Santa Barbara, San Diego, Riverside and Merced). Each UC is indicated according to student body population such that, the larger the circle, the more students attend the UC. Each UC marker additionally provides the UC website homepage URL in its pop-up. 

## Creating the map with Leaflet

```{r setup}
library(leaflet)
```

```{r mapping, message = FALSE}
set.seed(2020-08-17)

#data frame of latitude & longitude coordinates & population stats of UC's 
uc_data <- data.frame(name = c("UC Berkeley", "UC Davis", "UCLA", "UC Santa Cruz", "UC Irvine", "UC Santa Barbara", "UC San Diego", "UC Riverside", "UC Merced"),
                        pop = c(31352, 30872, 31577, 17792, 29736, 23070, 30285, 20581, 7881),
                        Lat = c(37.872111, 38.540681, 34.069024, 36.988059, 33.640477, 34.413910, 32.880105, 33.973759, 37.366057), 
                        Lng = c(-122.258518, -121.749119, -118.445095, -122.058209, -117.844243, -119.849001, -117.233949, -117.328182, -120.422525))

#custom popups according to UC homepage URL
uc_info <- c("https://www.berkeley.edu/",
             "https://www.ucdavis.edu/",
             "https://www.ucla.edu/",
             "https://www.ucsc.edu/",
             "https://uci.edu/",
             "https://www.ucsb.edu/",
             "https://ucsd.edu/",
             "https://www.ucr.edu/",
             "https://www.ucmerced.edu/")

#piping customization into leaflet map
uc_map <- uc_data %>% 
    leaflet() %>%
    addTiles() %>%
    addMarkers(popup = uc_info) %>%
    addCircles(weight = 1, radius = sqrt(uc_data$pop) * 150)
```

## Printing custom Leaflet map

```{r final, echo = FALSE}
uc_map
```