---
title: "Reproducible Research Project 1"
output: rmarkdown::github_document
  
---
##### Johns Hopkins Data Science Specialization : Course 5 : Week 2 : Project 1

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

## Introduction
This assignment makes use of data from a personal activity monitoring device, the specifics of which can be found in the project's README file. The aim of the project is create reproducible research through use of publicly available and interpretable analytic code, analytic data, well-documented analysis steps via markdown file, and the standard means of the sample distribution. Specifically, we will look at the steps taken by an individual over the course of 2 months in 5 minute intervals, the exact analyses of which are as follows: 

## Loading and preprocessing the data

First off, we'll load the 'tidyverse' & 'gridExtra' packages to make sure we have all the functionality we'll need for this particular project.

```{r libraries, echo = FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
```

Next, we'll load the data from the forked repository (found [HERE]('http://github.com/rdpeng/RepData_PeerAssessment1'). See README for more details regarding the data) into our local environment. 

```{r data}
zip <- unzip('activity.zip')
df <- read.csv('activity.csv')
head(df) #checking if data has been correctly loaded
```

This particular dataset doesn't require a lot of preprocessing but we'll go ahead and convert the 'date' variable to the 'Date' class so we'll be able to use it later on.

```{r date}
df$date <- ymd(df$date)
```

## What is the mean total number of steps taken per day?

To be able to compute this analysis, we first have to group our data according to date and, once grouped into its respective days, find the total sum of steps taken each day. To do so, we'll take advantage of dplyr piping in the tidyverse package. 

```{r average steps 1}
avsteps <- df %>% group_by(date) %>%
    summarise(total_steps = sum(steps))
```

Next, we'll visually display the resulting steps data in a histogram via the ggplot2 package also found in the tidyverse package.

```{r histogram 1}
hist <- ggplot(avsteps, aes(x = total_steps))+
    geom_histogram(binwidth = 1500, color="lightgreen", fill="darkgreen") +
    labs(title = "Number of steps taken per day",
         y = "Frequency", x = "Total Steps Taken")
hist
```

For the final step in this analysis, we want to explicitly print the mean and median number of steps taken per day using the data previously grouped by day & graphed above. To do so, simply use the mean() & median() function.

```{r mean/median 1}

meansteps = mean(avsteps$total_steps, na.rm = TRUE)
mediansteps = median(avsteps$total_steps, na.rm = TRUE)

message('Mean total number of steps taken per day: ', round(meansteps, 2))
message('Median total number of steps taken per day: ', round(mediansteps, 2))
```

## What is the average daily activity pattern?

Next, we're going to take a look at the average number of steps taken per 5-minute interval across all provided days. First, aggregate the step data according to interval, then take the average value of each 

```{r timeseries data 1}
series <- aggregate(steps ~ interval, df, mean)
head(series)
```

To visualize the activity pattern across each day, create a time series plot via ggplot2 with each interval on the x-axis & the average number of steps on the y-axis.

```{r timeseries plot1}
plot1 <- ggplot(series) 
plot1 + geom_line(aes(x=interval, y=steps), color = "darkgreen") +
    labs(x = "Time Interval (in minutes)", 
         y = "Average number of steps taken per day", 
         title = "Average Daily Steps")
```

We're also curious about which of the 5-minute intervals contains the maximum number of steps on average across all of the days. We'll use the data aggregated above to do so via indexing.

```{r max steps}
series[series$steps == max(series$steps),]
```

## Imputing missing values

We've previously been using the data without removing any missing values (indicated by 'NA' in the dataset). Let's do so now to see how it affects our previous mean and median step calculations. First, let's double check and see how many NA values are in the dataset.

```{r NA ount}
sum(is.na(df))
```

That's quite a few and they all seem to be in the 'steps' column. We'll replace each NA with the median value of steps per 5-minute interval. The steps cover a pretty far range in a non-normal distribution, thus we will pick the median over the mean (median values tend to be less influenced by outliers) and trim the outer 10% of the data to get a more representative missing value replacement value that is less likely to further skew the data when added in.

```{r NA add}

#Calculating the median value of steps per 5 minute interval
newsteps <- df %>% group_by(interval) %>% 
    summarise(steps = median(steps, na.rm = TRUE, trim = 0.05)) 
newsteps <- as.data.frame(newsteps)

#Replacing all NA's with calculated median values
cleandf <- full_join(df, newsteps, by = "interval") %>% 
    mutate(steps = ifelse(is.na(steps.x), steps.y, steps.x)) %>% 
    select(date, interval, steps)

#checking cleaned dataframe
sum(is.na(cleandf))
```

Great! As you can see, all of the NA values have now been replaced. Let's check to see if this has any affect on our previous mean & median calculations. First, we'll visualize the average number of steps taken per day just as before but now use our clean dataframe as the input.

```{r hist 2}
#grouping the data according to date & finding the sum of steps per day 
avsteps2 <- cleandf %>% group_by(date) %>%
    summarise(total_steps = sum(steps))

#plotting a histogram of the calculated steps per day
hist2 <- ggplot(avsteps2, aes(x = total_steps))+
    geom_histogram(binwidth = 1500, color="darkgreen", fill="lightpink") +
    labs(title = "Number of steps taken per day (NA values removed)",
         y = "Frequency", x = "Total Steps Taken")
hist2
```

It doesn't look like the distribution has changed much but let's compare the plots side-by-side just to be sure.

```{r hists}
grid.arrange(hist, hist2, ncol=2)
```

Awesome, the histograms look nearly identical. Let's calculate the mean and median steps per day with the cleaned data and compare to our previous calculations with the NA's still included to double check. 

```{r}
meansteps2 = mean(avsteps2$total_steps, na.rm = TRUE)
mediansteps2 = median(avsteps2$total_steps, na.rm = TRUE)

message('Mean total number of steps taken per day: ', round(meansteps2, 2))
message('Median total number of steps taken per day: ', round(mediansteps2, 2))

#creating a table to directly compare values 
values <- matrix(c(meansteps, mediansteps, mediansteps, mediansteps2), ncol = 2, byrow = T)
colnames(values) <- c("Mean of Steps Taken", "Median of Steps Taken")
rownames(values) <- c("Dataset with NA values", "Dataset without NA values")
values <- as.table(values)
values
```

The replacement of the NA values with the median interval values indeed has not changed the distribution much!

## Are there differences in activity patterns between weekdays and weekends?

Lastly, let's take a look at the patterns according to weekdays and weekends. To do so, we'll first create a new variable in our cleaned dataframe called 'daycategories' that groups each date into either 'Weekend' or 'Weekday' according to what day of the week the measurements were taken on.

```{r factor variable}
#using weekdays() to change year-month-date into weekdays
cleandf$days <- weekdays(cleandf$date)

#using new 'days' variable to create a factored variable according to weekend or weekday categorization
cleandf$daycategories <- ifelse(cleandf$days %in% c('Saturday', 'Sunday'), 'Weekend', 'Weekday')
cleandf$daycategories <- factor(cleandf$daycategories)
```

Next, we'll group our data according to interval and day category (weekend or weekday) to find the average number of steps taken during each. 

```{r series data 2}
series2 <- cleandf %>% group_by(interval, daycategories) %>%
    summarise(average_steps = mean(steps))
```

Finally, to visually compare the activity patterns of weekdays and weekends, we'll create another time series plot using a ggplot2 panel. 

```{r timeseries 2}
plot2 <- ggplot(series2) 
plot2 + geom_line(aes(x=interval, y=average_steps), color = "darkgreen") +
    facet_grid(daycategories ~ .) +
    labs(x = "Time Interval (in minutes)", 
         y = "Average number of steps taken per day", 
         title = "Average Daily Steps")
```


And we've reached the end of the project! Well done!